---
title: "ETL step 1: create the CDM and Synthea tables"
subtitle: "Notebook 1. Source script: etl/01_load_vocab.R"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include = FALSE}
# The notebook lives in notebooks/, the project files live one level up. Pointing knitr at the repository root
# means every relative path below (".Renviron", "results/", "seer/", "figures/") works exactly as in the .R scripts.
knitr::opts_knit$set(root.dir = normalizePath(".."))
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, comment = "#>")
run_etl <- FALSE   # TRUE = actually (re)create the database tables. Leave FALSE to just read the notebook.
```

## What this notebook does

Creates the empty OMOP CDM 5.4 tables (schema `cdm`) and the Synthea native tables (schema `native`) in PostgreSQL, using ETLSyntheaBuilder. Optionally loads the Athena
vocabularies from CSV. The vocabulary load is slow and memory-hungry in R, so the project loads them with `etl/01b_load_vocab_copy.sh` instead, and this step is mainly for
table creation.

**These cells are switched off by default** (`run_etl <- FALSE` in the setup chunk), because they change the database. Run them only when you really want to (re)create the tables.
To run: set `run_etl <- TRUE`, make sure `.Renviron` has the `PG_*`, `JDBC_DIR` and (optional) `VOCAB_DIR` settings, and run the chunks in order.

The connection settings live in `etl/00_connect.R`, which this notebook sources.

## Connect and create the tables

`source()` loads the shared connection details (`cd`, schema names, versions). The two `Create...Tables` calls build the empty schemas. Set `LOAD_VOCAB_IN_R=1` in `.Renviron` to also load vocabularies from `VOCAB_DIR`.

```{r, eval = run_etl}
source("etl/00_connect.R")
library(ETLSyntheaBuilder)
CreateCDMTables(cd, cdm_schema, cdm_version)
CreateSyntheaTables(cd, synthea_schema, synthea_version)
if (identical(Sys.getenv("LOAD_VOCAB_IN_R"), "1")) LoadVocabFromCsv(cd, cdm_schema, Sys.getenv("VOCAB_DIR", "vocab"))
cat("TABLES CREATED\n")
```
