---
title: "ETL step 2: map Synthea to OMOP CDM 5.4"
subtitle: "Notebook 2. Source script: etl/02_etl_synthea.R"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include = FALSE}
# The notebook lives in notebooks/, the project files live one level up. Pointing knitr at the repository root
# means every relative path below (".Renviron", "results/", "seer/", "figures/") works exactly as in the .R scripts.
knitr::opts_knit$set(root.dir = normalizePath(".."))
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, comment = "#>")
run_etl <- FALSE   # TRUE = actually run the ETL against the database (this TRUNCATES the clinical CDM tables first).
```

## What this notebook does

Maps the Synthea tables in schema `native` to OMOP CDM 5.4 in schema `cdm`, using the ETL-Synthea SQL, run **in stages**.

**Why in stages.** The package function `LoadEventTables()` runs about 20 `INSERT ... SELECT` steps back to back. A first attempt spent more than six hours in `insert_observation`
(cause not established; see `docs/provenance.md`). Here the same ETL-Synthea SQL is rendered with `sqlOnly = TRUE` and executed one step at a time, with an index on `final_visit_ids` and
`ANALYZE` between steps. The event load then finishes in about 10 minutes, without `drug_era`.

**Run `etl/02a_load_native_copy.sh` first** (it loads the Synthea CSVs into `native`).

**These cells are switched off by default** (`run_etl <- FALSE`). The first step empties the clinical CDM tables, so only run them when you intend to reload.
Options (set as environment variables in `.Renviron`):

- `ETL_SKIP_STEPS`: comma-separated step files to skip. Default `insert_drug_era.sql`: the OHDSI `drug_era` query ran for over an hour without finishing on about 500,000 exposures,
  and this project derives lines of therapy from `drug_exposure`, so `cdm.drug_era` stays empty. Set `none` to run it.
- `ETL_RESUME_FROM`: an event step file name (for example `insert_device_exposure.sql`) to resume an interrupted run.

## Connect and set up helpers

`run()` executes SQL, `q()` queries, `stamp()` prints a timestamped progress line.

```{r, eval = run_etl}
source("etl/00_connect.R")
library(ETLSyntheaBuilder)
cdm <- cdm_schema; nat <- synthea_schema
con <- connect(cd)
run <- function(sql) executeSql(con, sql, progressBar = FALSE, reportOverallTime = FALSE)
q <- function(sql) querySql(con, sql)
stamp <- function(...) cat(format(Sys.time(), "%H:%M:%S"), ..., "\n")
```

## Render the ETL-Synthea SQL

`render_steps()` asks ETL-Synthea for its SQL files without running them, and returns the file paths in execution order. `event_order` is the known order of the event-table steps.

```{r, eval = run_etl}
render_steps <- function(fun, ..., order = NULL) {   # render ETL-Synthea steps to SQL files, return them in execution order
  tmp <- tempfile("etlsql"); dir.create(tmp); owd <- setwd(tmp); on.exit(setwd(owd))
  out <- capture.output(fun(..., sqlOnly = TRUE))
  files <- list.files(file.path(tmp, "output"), pattern = "[.]sql$")
  if (is.null(order)) {   # CreateMapAndRollupTables prints "Saving to output/<file>" in execution order
    files <- sub("^Saving to output/", "", grep("^Saving to output/.*[.]sql$", out, value = TRUE))
  } else {                # LoadEventTables prints the SQL text instead of the name, so use the known step order
    files <- c(intersect(order, files), setdiff(files, order))
  }
  stopifnot(length(files) > 0)
  setNames(file.path(tmp, "output", files), files)
}
event_order <- c("insert_location.sql", "insert_care_site.sql", "insert_person.sql", "insert_observation_period.sql",
                 "insert_provider.sql", "insert_visit_occurrence.sql", "insert_visit_detail.sql", "insert_condition_occurrence.sql",
                 "insert_observation.sql", "insert_measurement.sql", "insert_procedure_occurrence.sql", "insert_drug_exposure.sql",
                 "insert_condition_era.sql", "insert_drug_era.sql", "insert_cdm_source.sql", "insert_device_exposure.sql",
                 "insert_death.sql", "insert_payer_plan_period.sql", "insert_cost_v270.sql", "insert_cost_v300.sql")
```

## Options

Reads `ETL_SKIP_STEPS` and `ETL_RESUME_FROM`, and defines a helper that checks whether a table exists.

```{r, eval = run_etl}
# Options (environment variables):
#   ETL_SKIP_STEPS  comma-separated ETL-Synthea step files to skip. Default: insert_drug_era.sql. The OHDSI drug_era query
#                   (recursive CTE over drug_exposure) ran over an hour on ~500k exposures in PostgreSQL without finishing;
#                   this project derives lines of therapy from drug_exposure, so cdm.drug_era is left empty. Set to "none" to run it.
#   ETL_RESUME_FROM an event step file name (e.g. insert_device_exposure.sql): skip the truncate, maps and earlier event steps
#                   and continue from that step, for resuming an interrupted run.
skip_steps  <- setdiff(strsplit(Sys.getenv("ETL_SKIP_STEPS", "insert_drug_era.sql"), ",")[[1]], "none")
resume_from <- Sys.getenv("ETL_RESUME_FROM", "")
exists_tbl <- function(schema, tbl) nrow(q(sprintf("select 1 from information_schema.tables where table_schema='%s' and table_name='%s'", schema, tbl))) > 0
```

## Steps 0 and 1: empty the clinical tables, build the maps

Empties the clinical CDM tables (vocabulary tables are untouched), builds the source-to-standard and source-to-source vocabulary maps once, builds the visit roll-ups, adds indexes and runs `ANALYZE`.

```{r, eval = run_etl}
# 0. fresh load: empty the clinical CDM tables (vocabulary tables are untouched)
clinical <- c("person", "observation_period", "visit_occurrence", "visit_detail", "condition_occurrence", "drug_exposure",
              "procedure_occurrence", "device_exposure", "measurement", "observation", "death", "note", "note_nlp", "specimen",
              "fact_relationship", "location", "care_site", "provider", "payer_plan_period", "cost", "drug_era", "dose_era",
              "condition_era", "episode", "episode_event", "metadata", "cdm_source")
if (!nzchar(resume_from)) {
for (t in clinical) if (exists_tbl(cdm, t)) run(sprintf("truncate table %s.%s", cdm, t))
stamp("clinical CDM tables emptied")

# 1. source-to-standard maps and visit roll-ups (the two vocabulary maps depend only on the vocabulary: build once)
vocab_maps <- c("create_source_to_standard_vocab_map.sql", "create_source_to_source_vocab_map.sql")
have_maps <- exists_tbl(cdm, "source_to_standard_vocab_map") && exists_tbl(cdm, "source_to_source_vocab_map") &&
  isTRUE(as.numeric(q(sprintf("select count(*) from %s.source_to_source_vocab_map", cdm))[[1]]) > 0)
steps <- render_steps(CreateMapAndRollupTables, cd, cdm, nat, cdm_version, synthea_version)
for (f in names(steps)) {
  if (have_maps && f %in% vocab_maps) { stamp("skip (already built)", f); next }
  stamp("running", f); run(SqlRender::readSql(steps[[f]]))
}
CreateExtraIndices(cd, cdm, nat, synthea_version)   # signature: (connectionDetails, cdmSchema, syntheaSchema, ...)
run(sprintf("create index if not exists idx_fvi_encounter on %s.final_visit_ids (encounter_id)", cdm))
for (t in q(sprintf("select table_name from information_schema.tables where table_schema='%s'", nat))[[1]]) run(sprintf("analyze %s.%s", nat, t))
for (t in c("final_visit_ids", "all_visits", "assign_all_visit_ids", "source_to_standard_vocab_map", "source_to_source_vocab_map")) run(sprintf("analyze %s.%s", cdm, t))
stamp("maps, indexes and statistics ready")
} else stamp("resuming at", resume_from, "(truncate, maps and earlier steps skipped)")
```

## Step 2: load the event tables one at a time

Runs each event step (person, visit, condition, drug, measurement and so on), running `ANALYZE` after each target table. The final table prints the row counts.

```{r, eval = run_etl}
# 2. event tables, one step at a time, with ANALYZE after each target table
steps <- render_steps(LoadEventTables, cd, cdm, nat, cdm_version, synthea_version, order = event_order)
started <- !nzchar(resume_from)
for (f in names(steps)) {
  if (!started && f == resume_from) started <- TRUE
  if (!started) next
  if (f %in% skip_steps) { stamp("skipped by ETL_SKIP_STEPS:", f); next }
  stamp("running", f); run(SqlRender::readSql(steps[[f]]))
  tbl <- sub("^insert_", "", sub("[.]sql$", "", f))
  if (exists_tbl(cdm, tbl)) run(sprintf("analyze %s.%s", cdm, tbl))
}
stamp("event tables loaded")
cnt <- q(sprintf("select relname, n_live_tup from pg_stat_user_tables where schemaname='%s' and n_live_tup>0 and relname in ('%s') order by 1", cdm, paste(clinical, collapse = "','")))
print(cnt)
cat("ETL DONE\n")
```

## Close the connection

```{r disconnect, eval = run_etl}
disconnect(con)
```

## Session information

The packages and versions this notebook ran with.

```{r sessioninfo}
sessionInfo()
```
