---
title: "Restricted mean survival time to 60 months: Synthea vs SEER"
subtitle: "Notebook 4. Source script: analysis/rmst.R"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include = FALSE}
# The notebook lives in notebooks/, the project files live one level up. Pointing knitr at the repository root
# means every relative path below (".Renviron", "results/", "seer/", "figures/") works exactly as in the .R scripts.
knitr::opts_knit$set(root.dir = normalizePath(".."))
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, comment = "#>")
has_seer_export <- file.exists(file.path(normalizePath(".."), "seer", "nsclc_survival_by_stage_2010-2015.csv"))
suppressPackageStartupMessages(library(ggplot2))
```

## What this notebook does

Restricted mean survival time (RMST) is the **area under the survival curve** up to a fixed time, here 60 months. It summarises a whole curve in one number
and does not need the proportional-hazards assumption.

- **Synthea:** RMST from the Kaplan-Meier fit (`summary(survfit, rmean = 60)`) with a normal 95% interval. The notebook also integrates the step function by hand and
  checks that the two agree.
- **SEER:** the area under the monthly observed-survival curve (trapezoid rule over 60 one-month intervals, survival 100% at month 0).

**Important:** the SEER monthly life-table export (`seer/nsclc_survival_by_stage_2010-2015.csv`) is **not in the repository** (SEER's data use agreement does not allow
publishing small monthly counts). It is on your machine only. The notebook reads just its cumulative-survival column, and only the resulting RMST values are written.
If the file is missing, the SEER part is skipped and the notebook says so.

**Output:** `results/rmst_synthea_vs_seer.csv`.

## Setup

Packages and constants. `TAU` is the 60-month horizon.

```{r}
suppressPackageStartupMessages({ library(DBI); library(RPostgres); library(survival); library(dplyr); library(readr); library(stringr) })
options(width = 120, digits = 5)
TAU <- 60
STAGES <- c("I", "II", "III", "IV")
dir.create("results", showWarnings = FALSE)
```

## Synthea: load the cohort

Main cohort and the 2010-2015 sensitivity cohort, straight from the dbt mart.

```{r}
if (!nzchar(Sys.getenv("PG_HOST"))) readRenviron(".Renviron")
con <- dbConnect(Postgres(), host = Sys.getenv("PG_HOST"), port = as.integer(Sys.getenv("PG_PORT")),
                 dbname = Sys.getenv("PG_DB"), user = Sys.getenv("PG_USER"), password = Sys.getenv("PG_PASSWORD"))
mart <- dbGetQuery(con, "select * from dbt.mart_nsclc_cohort")
dbDisconnect(con)
mart <- mart |> mutate(stage = factor(stage, levels = STAGES), followup_months = as.numeric(followup_months), event = as.integer(event))
cohorts <- list(main = mart, sensitivity = filter(mart, dx_2010_2015))
```

## Synthea: RMST from the Kaplan-Meier fit

The `stopifnot` inside checks the library's RMST against the area computed by hand from the KM step function in every stratum.

```{r}
syn <- bind_rows(lapply(names(cohorts), function(nm) {
  d <- cohorts[[nm]]
  fit <- survfit(Surv(followup_months, event) ~ stage, data = d)
  tb <- summary(fit, rmean = TAU)$table
  est <- tb[, grep("^[*]?rmean$", colnames(tb))]; se <- tb[, grep("se\\(rmean\\)", colnames(tb))]
  # independent check: area under the KM step function, integrated by hand
  chk <- vapply(STAGES, function(st) {
    s <- summary(fit[paste0("stage=", st)], censored = TRUE); t <- c(0, s$time[s$time < TAU], TAU); v <- c(1, s$surv[s$time < TAU])
    sum(diff(t) * v)
  }, numeric(1))
  stopifnot(all(abs(chk - est) < 1e-6))
  tibble(cohort = nm, stage = sub("stage=", "", rownames(tb)), n = tb[, "records"], synthea_rmst = est, se = se,
         lcl = pmax(est - 1.96 * se, 0), ucl = pmin(est + 1.96 * se, TAU))
}))
```

## SEER: area under the observed-survival curve

Requires the local SEER*Stat export described above. Only the `Observed Survival (Cum)` column is used.

```{r, eval = has_seer_export}
f <- "seer/nsclc_survival_by_stage_2010-2015.csv"
if (!file.exists(f)) stop("The local SEER*Stat monthly export ", f, " is needed (it is not in the repository).")
raw <- read_lines(f)
starts <- which(str_starts(raw, '"Page type"'))
life <- read_csv(I(raw[starts[2]:length(raw)]), show_col_types = FALSE) |>
  rename(stage_code = 2, month = Interval, surv = `Observed Survival (Cum)`) |>
  mutate(stage = c("0" = "I", "1" = "II", "2" = "III", "3" = "IV", "4" = "Unknown")[as.character(stage_code)],
         surv = parse_number(as.character(surv)) / 100) |>
  filter(stage %in% STAGES, month <= TAU) |> select(stage, month, surv)
seer <- bind_rows(lapply(STAGES, function(st) {
  s <- life |> filter(stage == st) |> arrange(month)
  stopifnot(identical(as.integer(s$month), 1:TAU))
  s_all <- c(1, s$surv)                                   # survival at months 0, 1, ..., 60
  tibble(stage = st, seer_rmst = sum((head(s_all, -1) + tail(s_all, -1)) / 2))   # trapezoid over 60 one-month intervals
}))
```

## Comparison table

RMST in months for both sources and the difference (Synthea minus SEER). Stage I is close (about 42.9 vs 43.5 months) even though the curves have very different shapes.

```{r, eval = has_seer_export}
out <- syn |> left_join(seer, by = "stage") |>
  mutate(stage = factor(stage, levels = STAGES), difference_months = synthea_rmst - seer_rmst,
         synthea_rmst_ci = sprintf("%.1f (%.1f-%.1f)", synthea_rmst, lcl, ucl)) |>
  arrange(factor(cohort, levels = c("main", "sensitivity")), stage) |>
  transmute(cohort, stage, synthea_n = n, synthea_rmst_months = round(synthea_rmst, 2), synthea_ci_lo = round(lcl, 2), synthea_ci_hi = round(ucl, 2),
            synthea_rmst_ci, seer_rmst_months = round(seer_rmst, 2), difference_months = round(difference_months, 2))
write_csv(out, "results/rmst_synthea_vs_seer.csv")
cat(sprintf("RMST to %d months (months). Synthea: Kaplan-Meier, normal 95%% CI. SEER: area under the monthly observed-survival curve (trapezoid). difference = Synthea - SEER.\n\n", TAU))
print(as.data.frame(out |> select(cohort, stage, synthea_n, synthea_rmst_ci, seer_rmst_months, difference_months)), row.names = FALSE)
cat("\nCheck: the Synthea RMST equals the hand-integrated KM area in every stratum (asserted in the script).\n")
```

## Chart: RMST by stage, Synthea vs SEER

RMST to 60 months (the maximum possible is 60). Stage I looks matched; stages II-IV are far apart. Error bars are the Synthea 95% interval.

```{r chart-rmst-bars, eval = has_seer_export, fig.width = 8, fig.height = 4.5}
rm_main <- out |> filter(cohort == "main") |> mutate(stage = factor(stage, STAGES))
bars <- bind_rows(
  rm_main |> transmute(stage, source = "Synthea", rmst = synthea_rmst_months, lo = synthea_ci_lo, hi = synthea_ci_hi),
  rm_main |> transmute(stage, source = "SEER 17", rmst = seer_rmst_months, lo = NA_real_, hi = NA_real_))
ggplot(bars, aes(stage, rmst, fill = source)) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  geom_errorbar(aes(ymin = lo, ymax = hi), position = position_dodge(0.8), width = 0.25, na.rm = TRUE) +
  geom_text(aes(y = 1.5, label = sprintf("%.1f", rmst)), position = position_dodge(0.8), vjust = 0, colour = "white", size = 3.6) +
  scale_fill_manual(NULL, values = c("Synthea" = "#1f5fa8", "SEER 17" = "#d1495b")) +
  scale_y_continuous(limits = c(0, TAU), breaks = seq(0, TAU, 12)) +
  labs(x = "Stage", y = "RMST to 60 months (months)") +
  theme_minimal() + theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

## Chart: same area, different shape (stage I)

RMST is the area under the curve. For stage I the two areas are almost equal, even though the synthetic curve is above SEER early and far below it late. The areas are shaded; the numbers in the title come from the table above.

```{r chart-rmst-area, eval = has_seer_export, fig.width = 8, fig.height = 4.5}
d1 <- filter(cohorts$main, stage == "I")
fit1 <- survfit(Surv(followup_months, event) ~ 1, data = d1)
grid <- seq(0, TAU, by = 0.1)
syn_curve <- tibble(t = grid, S = summary(fit1, times = grid, extend = TRUE)$surv)
s1 <- life |> filter(stage == "I") |> arrange(month)
seer_curve <- tibble(t = c(0, s1$month), S = c(1, s1$surv))
r1 <- out |> filter(cohort == "main", stage == "I")
ggplot() +
  geom_area(data = syn_curve, aes(t, S), fill = "#1f5fa8", alpha = 0.25) +
  geom_line(data = syn_curve, aes(t, S, colour = "Synthea"), linewidth = 1) +
  geom_area(data = seer_curve, aes(t, S), fill = "#d1495b", alpha = 0.18) +
  geom_line(data = seer_curve, aes(t, S, colour = "SEER 17"), linewidth = 1) +
  scale_colour_manual(NULL, values = c("Synthea" = "#1f5fa8", "SEER 17" = "#d1495b")) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Months since diagnosis", y = "Survival",
       title = sprintf("Stage I: RMST %.1f months (Synthea) vs %.1f months (SEER)", r1$synthea_rmst_months, r1$seer_rmst_months),
       subtitle = "Shaded area under each curve to 60 months") +
  theme_minimal() + theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

## Session information

The packages and versions this notebook ran with.

```{r sessioninfo}
sessionInfo()
```
