---
title: "SEER age by sex by stage, demographics comparison, and reweighting"
subtitle: "Notebook 6. Source script: analysis/seer_age_sex_reweighting.R"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include = FALSE}
# The notebook lives in notebooks/, the project files live one level up. Pointing knitr at the repository root
# means every relative path below (".Renviron", "results/", "seer/", "figures/") works exactly as in the .R scripts.
knitr::opts_knit$set(root.dir = normalizePath(".."))
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, comment = "#>")
has_seer_export <- file.exists(file.path(normalizePath(".."), "seer", "nsclc_age_sex_by_stage_2010-2015.csv"))
knitr::opts_chunk$set(eval = has_seer_export)
suppressPackageStartupMessages(library(ggplot2))
```

## What this notebook does

1. Reads **SEER's age by sex by stage counts** (Cumulative Summary only) from the local SEER*Stat export and writes a small summary table.
2. Compares the **demographics** of the synthetic cohort with SEER (sex and age bands).
3. Runs a **direct-standardisation sensitivity analysis**: the synthetic patients are reweighted so that, within each stage, their age and sex mix matches SEER's, and
   the Kaplan-Meier curves are redrawn. If demographics explained the gap to SEER, the curves would move. They do not.

**Important:** the SEER*Stat export `seer/nsclc_age_sex_by_stage_2010-2015.csv` is **not in the repository**. Only its Cumulative Summary (lines 1-316, patient counts) is read.
The life tables that follow contain small monthly counts that must never be published, so they are never read or printed. Every count written to the repository is asserted to be
at least 16, per the SEER data use agreement. If the export is missing, this notebook skips all steps and says so.

**Outputs:** `seer/nsclc_age_sex_by_stage_summary.csv`, `results/demographics_synthea_vs_seer.csv`, `results/km_synthea_vs_seer_reweighted.csv`.

## Setup

Packages, the four stages, the four age bands and the time points (months 12-60).

```{r, eval = TRUE}
suppressPackageStartupMessages({ library(DBI); library(RPostgres); library(survival); library(dplyr); library(tidyr); library(readr) })
options(width = 120, digits = 5)
dir.create("results", showWarnings = FALSE)
STAGES <- c("I", "II", "III", "IV"); BANDS <- c("50-64", "65-74", "75-84", "85+")
TIMES <- c(12, 24, 36, 48, 60)
```

## 1. SEER counts: age by sex by stage

Only the first 316 lines of the export are read. The code checks that the table has the expected 315 rows, that both sexes equal male plus female in every cell, and that the smallest published count is at least 16.

```{r}
f <- "seer/nsclc_age_sex_by_stage_2010-2015.csv"
if (!file.exists(f)) stop("The local SEER*Stat export ", f, " is needed (it is not in the repository).")
summ <- read_csv(I(readLines(f, n = 316)), show_col_types = FALSE) |>      # lines 1-316 only
  transmute(stage_code = .data[["Stage I-IV (AJCC 7th grouped)"]], age_code = .data[["Age recode with <1 year olds and 90+"]],
            sex_code = Sex, n = N)
stopifnot(nrow(summ) == 315, all(summ$n[summ$age_code < 11 | summ$age_code == 20] == 0))
stage_names <- c("0" = "I", "1" = "II", "2" = "III", "3" = "IV", "4" = "Unknown")
band_of <- function(a) ifelse(a %in% 11:13, "50-64", ifelse(a %in% 14:15, "65-74", ifelse(a %in% 16:17, "75-84", ifelse(a %in% 18:19, "85+", NA))))
sex_names <- c("0" = "both", "1" = "male", "2" = "female")
cells <- summ |> filter(!is.na(band_of(age_code))) |>
  mutate(stage = stage_names[as.character(stage_code)], age_band = band_of(age_code), sex = sex_names[as.character(sex_code)]) |>
  group_by(stage, age_band, sex) |> summarise(n = sum(n), .groups = "drop")
# internal consistency: both = male + female in every cell
chk <- cells |> pivot_wider(names_from = sex, values_from = n); stopifnot(all(chk$both == chk$male + chk$female))
# add the all-ages rows, and the percentages
all_ages <- cells |> group_by(stage, sex) |> summarise(n = sum(n), age_band = "All 50+", .groups = "drop")
out <- bind_rows(cells, all_ages)
stage_tot <- out |> filter(age_band == "All 50+", sex == "both") |> select(stage, stage_total = n)
out <- out |> left_join(stage_tot, by = "stage") |>
  mutate(pct_of_stage = round(100 * n / stage_total, 2)) |> select(stage, age_band, sex, n, pct_of_stage) |>
  arrange(factor(stage, levels = c(STAGES, "Unknown")), factor(age_band, levels = c(BANDS, "All 50+")), factor(sex, levels = c("both", "male", "female")))
cat("SEER counts: totals by stage (both sexes):\n"); print(as.data.frame(stage_tot), row.names = FALSE)
cat(sprintf("Smallest count in the output file: %d (must be >= 16); cells: %d\n", min(out$n), nrow(out)))
stopifnot(min(out$n) >= 16)                                                # DUA: no count under 16 may be published
write_csv(out, "seer/nsclc_age_sex_by_stage_summary.csv")

seer_cell <- cells |> filter(stage %in% STAGES, sex != "both")           # male and female cells, I-IV
seer_stage_n <- stage_tot |> filter(stage %in% STAGES)
```

## 2. Synthea cohorts

The synthetic cohort with the same age bands. Synthea has no patients aged 85 or over, because the generator draws age at the reference date between 50 and 90.

```{r}
if (!nzchar(Sys.getenv("PG_HOST"))) readRenviron(".Renviron")
con <- dbConnect(Postgres(), host = Sys.getenv("PG_HOST"), port = as.integer(Sys.getenv("PG_PORT")),
                 dbname = Sys.getenv("PG_DB"), user = Sys.getenv("PG_USER"), password = Sys.getenv("PG_PASSWORD"))
mart <- dbGetQuery(con, "select * from dbt.mart_nsclc_cohort"); dbDisconnect(con)
mart <- mart |> mutate(followup_months = as.numeric(followup_months), event = as.integer(event), age_at_dx = as.numeric(age_at_dx),
  stage = factor(stage, levels = STAGES), sexl = ifelse(sex == "M", "male", "female"),
  age_band = cut(age_at_dx, c(50, 65, 75, 85, Inf), right = FALSE, labels = BANDS))
stopifnot(!anyNA(mart$age_band))
cohorts <- list(main = mart, sensitivity = filter(mart, dx_2010_2015))
```

## 3. Demographics, Synthea vs SEER

Percent male and percent in each age band, by stage and overall. The synthetic cohort is about 77% male (SEER 53%) and much younger.

```{r}
dem_one <- function(d_synthea, seer_cells, label) {
  syn_n <- nrow(d_synthea); sn <- sum(seer_cells$n[seer_cells$sex == "male"]) + sum(seer_cells$n[seer_cells$sex == "female"])
  syn_band <- table(factor(d_synthea$age_band, levels = BANDS)) / syn_n * 100
  seer_band <- sapply(BANDS, function(b) sum(seer_cells$n[seer_cells$age_band == b])) / sn * 100
  tibble(stage = label, synthea_n = syn_n, seer_n = sn,
         synthea_pct_male = 100 * mean(d_synthea$sexl == "male"), seer_pct_male = 100 * sum(seer_cells$n[seer_cells$sex == "male"]) / sn,
         synthea_pct_50_64 = syn_band[[1]], seer_pct_50_64 = seer_band[[1]], synthea_pct_65_74 = syn_band[[2]], seer_pct_65_74 = seer_band[[2]],
         synthea_pct_75_84 = syn_band[[3]], seer_pct_75_84 = seer_band[[3]], synthea_pct_85plus = syn_band[[4]], seer_pct_85plus = seer_band[[4]])
}
dem <- bind_rows(lapply(names(cohorts), function(nm) {
  d <- cohorts[[nm]]
  bind_rows(lapply(STAGES, function(st) dem_one(filter(d, stage == st), filter(seer_cell, stage == st), st)),
            dem_one(d, seer_cell, "All staged (I-IV)")) |> mutate(cohort = nm, .before = 1)
}))
dem_out <- dem |> mutate(diff_pct_male_pp = synthea_pct_male - seer_pct_male, diff_50_64_pp = synthea_pct_50_64 - seer_pct_50_64,
  diff_65_74_pp = synthea_pct_65_74 - seer_pct_65_74, diff_75_84_pp = synthea_pct_75_84 - seer_pct_75_84, diff_85plus_pp = synthea_pct_85plus - seer_pct_85plus) |>
  mutate(across(where(is.numeric) & !c(synthea_n, seer_n), ~ round(.x, 1)))
write_csv(dem_out, "results/demographics_synthea_vs_seer.csv")
cat("\n==== DEMOGRAPHICS, Synthea vs SEER (percent; SEER = diagnosed 2010-2015, age 50+, known stage) ====\n")
print(as.data.frame(dem_out |> select(cohort, stage, synthea_n, seer_n, synthea_pct_male, seer_pct_male, synthea_pct_50_64, seer_pct_50_64,
  synthea_pct_65_74, seer_pct_65_74, synthea_pct_75_84, seer_pct_75_84, synthea_pct_85plus, seer_pct_85plus)), row.names = FALSE)
```

## Chart: age and sex pyramid, Synthea vs SEER

Women to the left, men to the right, as a percentage of each cohort. The synthetic cohort is younger, more male, and has nobody aged 85 or over (the generator draws age at the reference date between 50 and 90).

```{r chart-pyramid, fig.width = 9, fig.height = 4}
pyr <- bind_rows(
  mart |> count(age_band, sex = sexl) |> mutate(source = "Synthea"),
  seer_cell |> group_by(age_band, sex) |> summarise(n = sum(n), .groups = "drop") |> mutate(source = "SEER 17")) |>
  mutate(age_band = factor(age_band, BANDS)) |>
  tidyr::complete(source, age_band, sex, fill = list(n = 0)) |>
  group_by(source) |> mutate(pct = 100 * n / sum(n)) |> ungroup()
ggplot(pyr, aes(age_band, ifelse(sex == "female", -pct, pct), fill = source)) +
  geom_col(width = 0.8) + coord_flip() +
  facet_wrap(~ source) +
  scale_fill_manual(values = c("Synthea" = "#1f5fa8", "SEER 17" = "#d1495b"), guide = "none") +
  scale_y_continuous(labels = function(x) paste0(abs(x), "%")) +
  labs(x = "Age at diagnosis", y = "Percent of cohort (women left, men right)") +
  theme_minimal() + theme(panel.grid.minor = element_blank())
```

## Chart: percent male by stage

Share of men in each stage and overall. The synthetic cohort is about three-quarters male in every stage; SEER is about half.

```{r chart-pct-male, fig.width = 8, fig.height = 4}
sx <- dem_out |> filter(cohort == "main") |>
  transmute(stage = factor(stage, c(STAGES, "All staged (I-IV)")), Synthea = synthea_pct_male, `SEER 17` = seer_pct_male) |>
  tidyr::pivot_longer(-stage, names_to = "source", values_to = "pct_male")
ggplot(sx, aes(stage, pct_male, fill = source)) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  geom_text(aes(label = sprintf("%.0f%%", pct_male)), position = position_dodge(0.8), vjust = -0.4, size = 3.2) +
  scale_fill_manual(NULL, values = c("Synthea" = "#1f5fa8", "SEER 17" = "#d1495b")) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(x = NULL, y = "Percent male") +
  theme_minimal() + theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

## 4. Direct standardisation: building the weights

Each synthetic patient gets a weight so that within a stage the age-by-sex cells match SEER's. Cells where SEER has patients but Synthea has none cannot be weighted; they are dropped and SEER's distribution is renormalised. The diagnostics show how much weight that leaves, and the effective sample size.

```{r}
make_weights <- function(d, merge_75plus) {
  d <- d |> mutate(cell_band = if (merge_75plus) ifelse(age_band %in% c("75-84", "85+"), "75+", as.character(age_band)) else as.character(age_band))
  sc <- seer_cell |> mutate(cell_band = if (merge_75plus) ifelse(age_band %in% c("75-84", "85+"), "75+", age_band) else age_band) |>
    group_by(stage, cell_band, sex) |> summarise(seer_n = sum(n), .groups = "drop")
  syn_cells <- d |> count(stage, cell_band, sexl, name = "syn_n") |> rename(sex = sexl)
  cc <- full_join(sc, syn_cells, by = c("stage", "cell_band", "sex")) |> mutate(syn_n = coalesce(syn_n, 0L))
  cc <- cc |> group_by(stage) |> mutate(seer_prop = seer_n / sum(seer_n), supported = syn_n > 0,
                                        seer_prop_used = ifelse(supported, seer_n, 0) / sum(ifelse(supported, seer_n, 0)),
                                        coverage = sum(seer_prop[supported])) |> ungroup()
  d |> left_join(cc |> select(stage, cell_band, sex, syn_n, seer_prop_used), by = c("stage", "cell_band", "sexl" = "sex")) |>
    group_by(stage) |> mutate(w = seer_prop_used * n() / syn_n) |> ungroup() |> list(cells = cc)
}
wt_runs <- list("4 age bands" = FALSE, "3 age bands (75+ merged)" = TRUE)
weights_list <- lapply(wt_runs, function(m) make_weights(cohorts$main, m))
cat("\n==== REWEIGHTING: coverage and weight diagnostics (main cohort) ====\n")
cat("Cells where SEER has patients but Synthea has none cannot be weighted: they are dropped and SEER's distribution is renormalised over the supported cells.\n")
for (nm in names(weights_list)) {
  w <- weights_list[[nm]]; cc <- w$cells
  cat(sprintf("\n-- %s\n", nm))
  print(as.data.frame(cc |> filter(!supported) |> select(stage, cell_band, sex, seer_n, syn_n)), row.names = FALSE)
  d <- w[[1]]
  diag <- d |> group_by(stage) |> summarise(n = n(), seer_share_covered_pct = round(100 * first(cc$coverage[cc$stage == first(stage)]), 1),
    max_weight = round(max(w), 2), min_weight = round(min(w), 2), effective_n = round(sum(w)^2 / sum(w^2), 0), .groups = "drop")
  print(as.data.frame(diag), row.names = FALSE)
}
```

## Chart: the size of the weights

Each synthetic patient's weight (4 age bands). A weight of 1 means no change. Very large weights would make the result depend on a few patients; here the spread is modest, except where a cell is sparse.

```{r chart-weights, fig.width = 9, fig.height = 5}
w4 <- weights_list[["4 age bands"]][[1]]
ggplot(w4, aes(w)) +
  geom_histogram(bins = 30, fill = "#1f5fa8", colour = "white") +
  geom_vline(xintercept = 1, linetype = 2) +
  facet_wrap(~ stage, scales = "free", labeller = labeller(stage = function(x) paste("Stage", x))) +
  labs(x = "Weight", y = "Patients") +
  theme_minimal() + theme(panel.grid.minor = element_blank())
```

## 5. Weighted Kaplan-Meier at years 1-5

Reweighted curves use a robust variance. The key line is the last table: five-year stage I survival moves from 18.2% to between 17.8% and 19.0%, against 53.0% in SEER, and stages II-IV stay at 0%.

```{r}
km_at <- function(d, weighted) {
  fit <- if (weighted) survfit(Surv(followup_months, event) ~ stage, data = d, weights = w, id = person_id, robust = TRUE, conf.type = "log-log")
         else survfit(Surv(followup_months, event) ~ stage, data = d, conf.type = "log-log")
  unw <- survfit(Surv(followup_months, event) ~ stage, data = d)           # unweighted number at risk, for the S = 0 rule
  s <- summary(fit, times = TIMES, extend = TRUE); u <- summary(unw, times = TIMES, extend = TRUE)
  s_end <- vapply(split(fit$surv, rep(sub("stage=", "", names(fit$strata)), fit$strata)), function(v) tail(v, 1), numeric(1))
  tibble(stage = sub("stage=", "", as.character(s$strata)), months = s$time, n_risk = u$n.risk, S = s$surv, lo = s$lower, hi = s$upper) |>
    mutate(ends0 = unname(s_end[stage]) < 1e-9,
           across(c(S, lo, hi), ~ case_when(n_risk == 0 & ends0 ~ 0, n_risk == 0 ~ NA_real_, TRUE ~ .x)))
}
seer_km <- read_csv("seer/nsclc_survival_by_stage_summary.csv", show_col_types = FALSE) |>
  filter(months %in% TIMES) |> transmute(stage, months, seer_S = observed_survival_pct)
unw <- km_at(cohorts$main, FALSE)
rw <- bind_rows(lapply(names(weights_list), function(nm) km_at(weights_list[[nm]][[1]], TRUE) |> mutate(weighting = nm)))
fmt <- function(s, lo, hi) ifelse(is.na(s), NA_character_, sprintf("%.1f (%.1f-%.1f)", 100 * s, 100 * lo, 100 * hi))
res <- rw |> left_join(unw |> select(stage, months, u_S = S, u_lo = lo, u_hi = hi, u_n = n_risk), by = c("stage", "months")) |>
  left_join(seer_km, by = c("stage", "months")) |>
  mutate(stage = factor(stage, levels = STAGES), year = months / 12,
         unweighted_S_ci = fmt(u_S, u_lo, u_hi), reweighted_S_ci = fmt(S, lo, hi),
         change_pp = round(100 * (S - u_S), 1), diff_to_seer_unweighted_pp = round(100 * u_S - seer_S, 1), diff_to_seer_reweighted_pp = round(100 * S - seer_S, 1)) |>
  transmute(weighting, stage, year, synthea_n_at_risk_unweighted = u_n, unweighted_S_pct = round(100 * u_S, 1), unweighted_S_ci,
            reweighted_S_pct = round(100 * S, 1), reweighted_S_ci, change_pp, seer_S_pct = seer_S, diff_to_seer_unweighted_pp, diff_to_seer_reweighted_pp) |>
  arrange(factor(weighting, levels = names(wt_runs)), stage, year)
write_csv(res, "results/km_synthea_vs_seer_reweighted.csv", na = "")
cat("\n==== WEIGHTED KAPLAN-MEIER AT YEARS 1-5 (main cohort; S in %, 95% log-log CI from a robust variance) ====\n")
print(as.data.frame(res |> select(weighting, stage, year, synthea_n_at_risk_unweighted, unweighted_S_ci, reweighted_S_ci, change_pp, seer_S_pct)), row.names = FALSE)
cat("\n==== CHANGE IN 5-YEAR SURVIVAL AFTER REWEIGHTING (percentage points) ====\n")
print(as.data.frame(res |> filter(year == 5) |> select(weighting, stage, unweighted_S_pct, reweighted_S_pct, change_pp, seer_S_pct)), row.names = FALSE)
```

## Chart: stage I survival before and after reweighting

Stage I survival at years 1-5: unweighted (solid blue), reweighted to SEER's age and sex mix (light blue, two weighting schemes) and SEER (red). Reweighting barely moves the curve, so demographic differences do not explain the gap. Stages II-IV are zero either way.

```{r chart-reweighted, fig.width = 8, fig.height = 4.5}
s1 <- res |> filter(stage == "I")
unw1 <- s1 |> filter(weighting == first(weighting)) |> select(year, S = unweighted_S_pct)
ggplot(s1, aes(year)) +
  geom_line(data = unw1, aes(y = S, colour = "Synthea, unweighted"), linewidth = 1.1) +
  geom_line(aes(y = reweighted_S_pct, colour = "Synthea, reweighted", linetype = weighting), linewidth = 0.9) +
  geom_line(aes(y = seer_S_pct, colour = "SEER 17"), linewidth = 1.1) +
  geom_point(aes(y = seer_S_pct, colour = "SEER 17"), size = 2.8) +
  scale_colour_manual(NULL, values = c("Synthea, unweighted" = "#1f5fa8", "Synthea, reweighted" = "#6fa8e8", "SEER 17" = "#d1495b")) +
  scale_linetype_manual("Weighting", values = c("4 age bands" = "dashed", "3 age bands (75+ merged)" = "dotted")) +
  scale_y_continuous(limits = c(0, 100), labels = function(x) paste0(x, "%")) +
  labs(x = "Years since diagnosis", y = "Observed survival") +
  theme_minimal() + theme(legend.position = "bottom", legend.box = "vertical", panel.grid.minor = element_blank())
```

## Session information

The packages and versions this notebook ran with.

```{r sessioninfo}
sessionInfo()
```
