---
title: "Descriptive treatment patterns in the NSCLC cohort"
subtitle: "Notebook 7. Source script: analysis/treatment_patterns.R"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include = FALSE}
# The notebook lives in notebooks/, the project files live one level up. Pointing knitr at the repository root
# means every relative path below (".Renviron", "results/", "seer/", "figures/") works exactly as in the .R scripts.
knitr::opts_knit$set(root.dir = normalizePath(".."))
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, comment = "#>")
suppressPackageStartupMessages(library(ggplot2))
```

## What this notebook does

A descriptive look at what treatment the 1,347 main-cohort patients received, from `drug_exposure` and `procedure_occurrence`. No line-of-therapy logic is applied here
(that is built in dbt, in `int_lines_of_therapy`).

The finding that matters: **1,346 of 1,347 patients received the same first-line treatment**. Treatment does not vary by stage, so a first-line treatment comparison
(for example with inverse probability of treatment weighting) is not identifiable in this data.

This notebook only reads from the database; it writes no files.

## Setup and connection

Connection to PostgreSQL and a small helper that prints section headings.

```{r}
suppressPackageStartupMessages({ library(DBI); library(RPostgres); library(dplyr); library(tidyr) })
options(width = 120, digits = 4, dplyr.summarise.inform = FALSE)
if (!nzchar(Sys.getenv("PG_HOST"))) readRenviron(".Renviron")
con <- dbConnect(Postgres(), host = Sys.getenv("PG_HOST"), port = as.integer(Sys.getenv("PG_PORT")),
                 dbname = Sys.getenv("PG_DB"), user = Sys.getenv("PG_USER"), password = Sys.getenv("PG_PASSWORD"))
q <- function(sql) dbGetQuery(con, sql)
STAGES <- c("I", "II", "III", "IV")
hdr <- function(x) cat("\n====", x, "====\n")
```

## The cohort

The main cohort, as defined by the dbt mart.

```{r}
cohort <- q("select person_id, index_date, stage, sex, event from dbt.mart_nsclc_cohort") |> mutate(stage = factor(stage, STAGES))
N <- nrow(cohort)
```

## Anticancer drugs and radiation procedures

Antineoplastic and endocrine anticancer ingredients (ATC L01 and L02) and radiation or chemoradiation procedures received by cohort members. Treatment is counted from the NSCLC index date onward; earlier exposures are examined separately in section 0.

```{r}
anti <- q("
  select de.person_id, de.drug_exposure_start_date as d, ing.concept_name as ingredient
  from cdm.drug_exposure de
  join dbt.mart_nsclc_cohort m using (person_id)
  join cdm.concept_ancestor ci on ci.descendant_concept_id = de.drug_concept_id
  join cdm.concept ing on ing.concept_id = ci.ancestor_concept_id and ing.concept_class_id = 'Ingredient'
  where exists (select 1 from cdm.concept_ancestor ca2
                join cdm.concept atc on atc.concept_id = ca2.ancestor_concept_id and atc.vocabulary_id = 'ATC'
                     and (atc.concept_code like 'L01%' or atc.concept_code like 'L02%')
                where ca2.descendant_concept_id = ing.concept_id)") |>
  mutate(d = as.Date(d)) |> distinct()
rad <- q("
  select po.person_id, po.procedure_date as d, po.procedure_source_value as code, c.concept_name as proc
  from cdm.procedure_occurrence po join dbt.mart_nsclc_cohort m using (person_id)
  join cdm.concept c on c.concept_id = po.procedure_concept_id
  where po.procedure_source_value in ('703423002', '33195004', '367336001', '394894008')") |> mutate(d = as.Date(d))
all_ing <- q("select count(distinct ing.concept_id) n from cdm.drug_exposure de join dbt.mart_nsclc_cohort m using (person_id)
              join cdm.concept_ancestor ci on ci.descendant_concept_id = de.drug_concept_id
              join cdm.concept ing on ing.concept_id = ci.ancestor_concept_id and ing.concept_class_id = 'Ingredient'")$n |> as.integer()
CORE <- c("cisplatin", "paclitaxel")                       # the NSCLC-directed agents in the Synthea module
idx <- cohort |> transmute(person_id, index_date = as.Date(index_date))
anti_all <- anti; rad_all <- rad
anti <- anti |> inner_join(idx, by = "person_id") |> filter(d >= index_date) |> select(-index_date)   # treatment = on or after the NSCLC index date
rad  <- rad  |> inner_join(idx, by = "person_id") |> filter(d >= index_date) |> select(-index_date)
pre_anti <- anti_all |> inner_join(idx, by = "person_id") |> filter(d < index_date)
pre_rad  <- rad_all  |> inner_join(idx, by = "person_id") |> filter(d < index_date)
```

## 0. Events before the index date

Exposures dated before the NSCLC diagnosis are excluded from the treatment analysis. These come from earlier cancers or hormone preparations in the patient's synthetic history.

```{r}
hdr("0. EVENTS BEFORE THE NSCLC INDEX DATE (excluded from everything below: earlier cancers or hormone preparations)")
cat(sprintf("patients with any anticancer-ingredient exposure before the index date: %d (%.1f%%); with a chemotherapy/radiation procedure before the index date: %d (%.1f%%)
",
    n_distinct(pre_anti$person_id), 100 * n_distinct(pre_anti$person_id) / N, n_distinct(pre_rad$person_id), 100 * n_distinct(pre_rad$person_id) / N))
cat("pre-index procedures by code and how many days before the index date (median, range):
")
print(as.data.frame(pre_rad |> mutate(days_before = as.numeric(index_date - d)) |> group_by(proc) |>
  summarise(patients = n_distinct(person_id), median_days_before = median(days_before), min_days_before = min(days_before), max_days_before = max(days_before))), row.names = FALSE)
```

## 1. Ingredients

Which anticancer ingredients appear. Only cisplatin and paclitaxel are the NSCLC-directed agents in the Synthea module.

```{r}
hdr("1. INGREDIENTS")
cat(sprintf("Distinct drug ingredients of any kind received by the %d patients: %d (mostly vaccines, cardiovascular and other chronic-disease drugs).\n", N, all_ing))
cat("Antineoplastic or endocrine anticancer ingredients (ATC L01 / L02): patients, administrations (distinct person-date-ingredient)\n")
ing_tab <- anti |> group_by(ingredient) |> summarise(patients = n_distinct(person_id), administrations = n(), pct_of_cohort = round(100 * n_distinct(person_id) / N, 1)) |>
  arrange(desc(patients)); print(as.data.frame(ing_tab), row.names = FALSE)
```

## Chart: anticancer ingredients received

How many of the cohort received each antineoplastic or endocrine anticancer ingredient. Cisplatin and paclitaxel (the NSCLC-directed agents) are given to almost everyone; the other ingredients belong to other cancers in the patients' synthetic histories.

```{r chart-ingredients, fig.width = 7, fig.height = 3.5}
ggplot(ing_tab, aes(reorder(ingredient, patients), patients)) +
  geom_col(fill = "#1f5fa8") +
  geom_text(aes(label = patients), hjust = -0.15, size = 3.2) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = sprintf("Patients (of %d)", N)) +
  theme_minimal() + theme(panel.grid.minor = element_blank())
```

## Per-patient treatment summaries

One row per patient: first systemic treatment, number of administration dates, first chemoradiation procedure, and time to first treatment.

```{r}
core <- anti |> filter(ingredient %in% CORE)
per <- cohort |>
  left_join(core |> group_by(person_id) |> summarise(first_sys = min(d), last_sys = max(d), n_dates = n_distinct(d), n_core_ing = n_distinct(ingredient)), by = "person_id") |>
  left_join(anti |> group_by(person_id) |> summarise(first_anyanti = min(d), n_anti_ing = n_distinct(ingredient)), by = "person_id") |>
  left_join(rad |> filter(code == "703423002") |> group_by(person_id) |> summarise(first_cxrt = min(d), n_cxrt = n()), by = "person_id") |>
  left_join(rad |> filter(code == "33195004") |> group_by(person_id) |> summarise(n_ebrt = n()), by = "person_id") |>
  mutate(any_sys = !is.na(first_sys), any_rad = !is.na(first_cxrt) | !is.na(n_ebrt),
         days_to_sys = as.numeric(first_sys - index_date), days_to_rad = as.numeric(first_cxrt - index_date),
         first_tx = pmin(first_sys, first_cxrt, na.rm = TRUE), days_to_first_tx = as.numeric(first_tx - index_date),
         span_days = as.numeric(last_sys - first_sys))
```

## 2. Systemic treatment and radiation by stage

```{r}
hdr("2. SYSTEMIC TREATMENT AND RADIATION BY STAGE (core = cisplatin or paclitaxel; radiation = combined chemoradiation or external beam procedure)")
print(as.data.frame(per |> group_by(stage) |> summarise(n = n(), any_systemic_n = sum(any_sys), any_systemic_pct = round(100 * mean(any_sys), 1),
        any_radiation_n = sum(any_rad), any_radiation_pct = round(100 * mean(any_rad), 1),
        both_n = sum(any_sys & any_rad), neither_n = sum(!any_sys & !any_rad))), row.names = FALSE)
cat(sprintf("All: any systemic %d (%.1f%%), any radiation %d (%.1f%%), neither %d\n", sum(per$any_sys), 100 * mean(per$any_sys), sum(per$any_rad), 100 * mean(per$any_rad), sum(!per$any_sys & !per$any_rad)))
cat("Patients without a recorded systemic or radiation treatment:\n"); print(as.data.frame(per |> filter(!any_sys | !any_rad) |> select(person_id, stage, sex, any_sys, any_rad, n_cxrt, n_ebrt)), row.names = FALSE)
```

## 3. Time from diagnosis to first treatment

```{r}
hdr("3. TIME FROM DIAGNOSIS TO FIRST TREATMENT (days)")
tt <- function(x) c(n = sum(!is.na(x)), median = median(x, na.rm = TRUE), q1 = unname(quantile(x, .25, na.rm = TRUE)), q3 = unname(quantile(x, .75, na.rm = TRUE)), min = min(x, na.rm = TRUE), max = max(x, na.rm = TRUE))
cat("first systemic (cisplatin/paclitaxel) after the index date:\n"); print(round(rbind(All = tt(per$days_to_sys), sapply(split(per$days_to_sys, per$stage), tt) |> t()), 1))
cat("first treatment of any kind (systemic or combined chemoradiation procedure):\n"); print(round(rbind(All = tt(per$days_to_first_tx), sapply(split(per$days_to_first_tx, per$stage), tt) |> t()), 1))
cat(sprintf("Patients whose first treatment is dated BEFORE the index date: %d\n", sum(per$days_to_first_tx < 0, na.rm = TRUE)))
```

## Chart: days from diagnosis to first systemic treatment

Distribution of the delay between the NSCLC diagnosis and the first cisplatin or paclitaxel administration, by stage.

```{r chart-time-to-treatment, fig.width = 8, fig.height = 3.8}
ggplot(per |> filter(!is.na(days_to_sys)), aes(days_to_sys)) +
  geom_histogram(bins = 30, fill = "#1f5fa8", colour = "white") +
  facet_wrap(~ stage, nrow = 1, labeller = labeller(stage = function(x) paste("Stage", x))) +
  labs(x = "Days from diagnosis to first systemic treatment", y = "Patients") +
  theme_minimal() + theme(panel.grid.minor = element_blank())
```

## 4. Regimens per patient

A regimen is the set of anticancer ingredients given on the same date.

```{r}
hdr("4. REGIMENS PER PATIENT (regimen = the set of anticancer ingredients given on the same date)")
reg <- anti |> group_by(person_id, d) |> summarise(regimen = paste(sort(ingredient), collapse = " + "), .groups = "drop")
reg_pp <- reg |> group_by(person_id) |> summarise(n_regimens = n_distinct(regimen), regimens = paste(sort(unique(regimen)), collapse = " | "))
cat("distinct regimens per patient (all anticancer ingredients):\n"); print(table(factor(reg_pp$n_regimens)))
cat("most common regimen sets across patients:\n"); print(as.data.frame(reg_pp |> count(regimens, sort = TRUE) |> head(8)), row.names = FALSE)
reg_core <- reg |> mutate(regimen_core = vapply(strsplit(regimen, " [+] "), function(x) paste(sort(intersect(x, CORE)), collapse = " + "), "")) |> filter(nzchar(regimen_core))
cat("regimens containing only the core agents, by date: how often each combination is given on a date\n"); print(table(reg_core$regimen_core))
```

## 5. Gaps between administrations

The distribution of gaps between consecutive administration dates. This informs the 90-day gap rule used to start a new line of therapy.

```{r}
hdr("5. GAPS BETWEEN ADMINISTRATIONS (core agents; consecutive distinct administration dates within a patient)")
gaps <- core |> distinct(person_id, d) |> arrange(person_id, d) |> group_by(person_id) |> mutate(gap = as.numeric(d - lag(d))) |> ungroup() |> filter(!is.na(gap))
cat(sprintf("administration dates per patient: median %.0f (IQR %.0f-%.0f), max %d; days between first and last administration: median %.0f (IQR %.0f-%.0f), max %.0f\n",
    median(per$n_dates, na.rm = TRUE), quantile(per$n_dates, .25, na.rm = TRUE), quantile(per$n_dates, .75, na.rm = TRUE), max(per$n_dates, na.rm = TRUE),
    median(per$span_days, na.rm = TRUE), quantile(per$span_days, .25, na.rm = TRUE), quantile(per$span_days, .75, na.rm = TRUE), max(per$span_days, na.rm = TRUE)))
cat("gap lengths (days) between consecutive administrations:\n"); print(round(quantile(gaps$gap, c(0, .01, .05, .25, .5, .75, .95, .99, 1)), 1))
gp <- gaps |> group_by(person_id) |> summarise(max_gap = max(gap), any_gt28 = any(gap > 28), any_gt90 = any(gap > 90))
cat(sprintf("patients with any gap > 28 days: %d (%.1f%% of treated); any gap > 90 days: %d (%.1f%%); largest gap %.0f days\n",
    sum(gp$any_gt28), 100 * mean(gp$any_gt28), sum(gp$any_gt90), 100 * mean(gp$any_gt90), max(gp$max_gap)))
cat("gap histogram (days): "); print(table(cut(gaps$gap, c(-1, 0, 1, 7, 14, 28, 60, 90, 180, 365, Inf))))
```

## Chart: gaps between administrations

Days between consecutive administration dates (log scale). The dashed lines mark the 28-day and 90-day thresholds used in the line-of-therapy rules. Most gaps are short; very few exceed 90 days, which is why very few patients reach a second line.

```{r chart-gaps, fig.width = 8, fig.height = 3.8}
ggplot(gaps, aes(gap)) +
  geom_histogram(bins = 40, fill = "#1f5fa8", colour = "white") +
  geom_vline(xintercept = c(28, 90), linetype = 2) +
  annotate("text", x = c(28, 90), y = Inf, label = c("28 days", "90 days"), vjust = 1.5, hjust = -0.1, size = 3.3) +
  scale_x_log10() +
  labs(x = "Days between consecutive administrations (log scale)", y = "Gaps") +
  theme_minimal() + theme(panel.grid.minor = element_blank())
```

## 6. Line-2 candidates under the planned rules

How many patients would start a second line under a naive rule, and why restricting to the NSCLC-directed agents matters.

```{r}
hdr("6. LINE-2 CANDIDATES UNDER THE PLAN.md RULES (a gap > 90 days, or a new drug started > 28 days after first treatment)")
first_any  <- anti |> group_by(person_id) |> summarise(first_d = min(d))
first_core <- core |> group_by(person_id) |> summarise(first_core_d = min(d))
newdrug <- anti |> inner_join(first_any, by = "person_id") |> group_by(person_id, ingredient) |> summarise(first_ing = min(d), first_d = first(first_d), .groups = "drop") |>
  mutate(lag_days = as.numeric(first_ing - first_d)) |> filter(lag_days > 28)
newcore <- core |> inner_join(first_core, by = "person_id") |> group_by(person_id, ingredient) |> summarise(first_ing = min(d), first_core_d = first(first_core_d), .groups = "drop") |>
  mutate(lag_days = as.numeric(first_ing - first_core_d)) |> filter(lag_days > 28)
cand_new_all  <- unique(newdrug$person_id)
cand_new_core <- unique(newcore$person_id)
cand_gap      <- gp$person_id[gp$any_gt90]
cat(sprintf("naive rule, ANY anticancer/endocrine ingredient after the index date: new drug > 28 days after the first: %d patients (%.1f%%); gap > 90 days (core agents): %d (%.1f%%); either: %d (%.1f%%)
",
    length(cand_new_all), 100 * length(cand_new_all) / N, length(cand_gap), 100 * length(cand_gap) / N, length(union(cand_new_all, cand_gap)), 100 * length(union(cand_new_all, cand_gap)) / N))
cat(sprintf("restricted to the NSCLC-directed agents (cisplatin, paclitaxel), measured from the first cisplatin/paclitaxel date: new core drug > 28 days after the first: %d; gap > 90 days: %d; either: %d (%.1f%%)
",
    length(cand_new_core), length(cand_gap), length(union(cand_new_core, cand_gap)), 100 * length(union(cand_new_core, cand_gap)) / N))
cat("which ingredients create the 'new drug > 28 days' candidates:\n"); print(as.data.frame(newdrug |> count(ingredient, sort = TRUE) |> mutate(pct_of_cohort = round(100 * n / N, 1))), row.names = FALSE)
oth <- anti |> filter(!ingredient %in% CORE)
cat("\nthose non-core ingredients: sex of recipients and co-administered drugs (are they another cancer's regimen?)\n")
print(as.data.frame(oth |> inner_join(cohort |> select(person_id, sex), by = "person_id") |> group_by(ingredient) |> summarise(patients = n_distinct(person_id), pct_male = round(100 * mean(sex[!duplicated(person_id)] == "M"), 0),
        same_day_as_leuprolide = n_distinct(person_id[person_id %in% (anti$person_id[anti$ingredient == "leuprolide"])])) |> arrange(desc(patients)) |> head(8)), row.names = FALSE)
dl <- anti |> filter(ingredient == "docetaxel") |> inner_join(anti |> filter(ingredient == "leuprolide"), by = c("person_id", "d"))
cat(sprintf("docetaxel administrations on the same date as leuprolide: %d of %d docetaxel administrations; patients with both: %d of %d docetaxel patients\n",
    nrow(dl), sum(anti$ingredient == "docetaxel"), n_distinct(dl$person_id), n_distinct(anti$person_id[anti$ingredient == "docetaxel"])))
```

## 7. Is treatment type determined by stage?

First-line type by stage, with Cramer's V and a chi-square test. Treatment is essentially constant across stages.

```{r}
hdr("7. IS TREATMENT TYPE DETERMINED BY STAGE? (first-line type: chemoradiation if the combined procedure is within 28 days of the first systemic, else chemotherapy alone, else none)")
per <- per |> mutate(first_line = case_when(
  any_sys & !is.na(first_cxrt) & abs(as.numeric(first_cxrt - first_sys)) <= 28 ~ "chemotherapy + radiation",
  any_sys ~ "chemotherapy alone", !any_sys & any_rad ~ "radiation only", TRUE ~ "none recorded"))
tab <- table(Stage = per$stage, `First-line type` = per$first_line); print(tab)
cat("row percentages:\n"); print(round(100 * prop.table(tab, 1), 1))
cat(sprintf("Cramer's V (stage vs first-line type): %.3f; chi-square p = %s\n", {ct <- suppressWarnings(chisq.test(tab)); sqrt(ct$statistic / (sum(tab) * (min(dim(tab)) - 1)))},
    format.pval(suppressWarnings(chisq.test(tab))$p.value, digits = 3)))
cat(sprintf("patients in the largest first-line group: %d of %d (%.1f%%)\n", max(table(per$first_line)), N, 100 * max(table(per$first_line)) / N))
cat("treatment intensity, if everyone is treated alike: administrations (dates) per patient by stage\n")
print(as.data.frame(per |> group_by(stage) |> summarise(median_dates = median(n_dates, na.rm = TRUE), q1 = quantile(n_dates, .25, na.rm = TRUE), q3 = quantile(n_dates, .75, na.rm = TRUE),
        median_span_days = median(span_days, na.rm = TRUE), pct_died = round(100 * mean(event), 0))), row.names = FALSE)
dbDisconnect(con)
```

## Chart: first-line type by stage

Proportion of each stage in each first-line group. Almost the whole cohort sits in one group in every stage, so there is no treatment contrast to analyse.

```{r chart-first-line, fig.width = 7, fig.height = 4}
ggplot(per, aes(stage, fill = first_line)) +
  geom_bar(position = "fill", width = 0.7) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_viridis_d(name = "First line", end = 0.85) +
  labs(x = "Stage", y = "Share of patients") +
  theme_minimal() + theme(panel.grid.minor = element_blank())
```

## Session information

The packages and versions this notebook ran with.

```{r sessioninfo}
sessionInfo()
```
